Explainability and Adversarial Robustness for RNNs

### 2020-02-14 paper/paper.tex
Readded PDP and feature dropout table

We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e. forward or reverse path) and all TCP flags (0 if the flow is not TCP).
% \begin{itemize*}
% \item source port
% \item destination port
% \item protocol identifier
% \item packet sizes
% \item \glspl{iat} of packets
% \item packet directions (if packet goes there or back)
% \item all TCP flags (0 if the flow is not TCP)
% \end{itemize*}.
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behavior \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization and a train/test split of 2:1.

For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks. Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
\autoref{tab:performance_results} shows that our \gls{rnn}-based classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}. However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before the attack flows terminate.

\begin{table}
% Evaluated using
% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
% on Dec. 4th
\caption{Performance metrics per packet and per flow. MLP values from \cite{bachl_walling_2019} are presented for comparison.} \label{tab:performance_results}
\centering
\begin{tabular}{l r r r r r r} \toprule
& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
	&	Packet	&	Flow	& MLP &	Packet	&	Flow & MLP	\\	\midrule
Accuracy	&	99.1\%	&	99.7\%	&	99.8\%	&	99.5\%	&	98.3\%	&	98.9\%	\\
Precision	&	97.0\%	&	99.7\%	&	99.9\%	&	83.4\%	&	78.6\%	&	84.5\%	\\
Recall	&	97.8\%	&	99.1\%	&	99.2\%	&	87.6\%	&	72.6\%	&	82.9\%	\\
F1	&	97.4\%	&	99.4\%	&	99.5\%	&	85.5\%	&	75.5\%	&	83.7\%	\\
Youden	&	97.2\%	&	99.0\%	&	99.1\%	&	87.3\%	&	71.9\%	&	82.3\%	\\
\bottomrule
\end{tabular}
%\vspace*{-0.3cm}
\end{table}

@@

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{./plots/adv_comparison/tradeoff.pdf}
\caption{Performance of \gls{cw} with different $\kappa$ values.}
\label{fig:adv_cw}
\end{figure}

@@

\subsection{Feature Importance Metrics}
%\begin{table}
%\caption{Accuracy drop for:}
%\label{tab:feat_importance}
%\centering
%\subfloat[Random replacement method\label{fig:featRandom}
%]{
%\begin{tabular}{l r}
%\toprule
%Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
%Protocol & 0.207 \\
%Packet Length & 0.165 \\
%SYN Flag & 0.099 \\
%ACK Flag & 0.084 \\
%Direction & 0.073 \\
%Destination port & 0.071 \\
%Source port & 0.060 \\
%RST Flag & 0.057 \\
%PSH Flag & 0.056 \\
%Interarrival time & 0.024 \\
%FIN Flag & 0.012 \\
%URG Flag & 0 \\
%ECE Flag & 0 \\
%CWR Flag & 0 \\
%NS Flag & 0 \\
%\bottomrule
%\end{tabular}
%}{}
%\subfloat[Feature dropout\label{fig:featDropout}
%]{
%\begin{tabular}{l r}
%\toprule
%Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
%Destination port & 0.025 \\
%Source port & 0.003 \\
%RST Flag & 0.001 \\
%ACK Flag & 0.001 \\
%Protocol & 0.001 \\
%Packet Length & 0.001 \\
%Direction & 0.001 \\
%SYN Flag & 0 \\
%Interarrival time & 0 \\
%FIN Flag & 0 \\
%ECE Flag & 0 \\
%URG Flag & 0 \\
%CWR Flag & 0 \\
%NS Flag & 0 \\
%PSH Flag & 0 \\
%\bottomrule
%\end{tabular}
%}{}
%\end{table}
As a first step to understanding the neural network's decisions, we  estimate how important individual features are for the model's predictions.
When investigating an \gls{ml} classifier, it is natural to ask which inputs have a large influence on the classifier's prediction. We feel the need to distinguish metrics for this purpose based on their main aim:

@@

\subsubsection{Neural Network Weights}
In previous works~\cite{olden_accurate_2004}, a simple method for determining feature importance in neural networks has been summing up neural network weights leading from a certain input to the prediction outcome. The weights method can be considered for both feature importance and feature sensitivity, however, clearly, especially in the case of complex network architectures, this method is likely to provide wrong results. Hence, we provide results for the weights method mainly for comparison. Note that an LSTM cell alone has four weights leading from one input to an output.

@@

An important implementation detail is that for each feature we add another input which is 1 if the feature is suppressed and 0 otherwise. This is necessary for the neural network to be able to distinguish between a feature missing and a feature genuinely being zero. The overall outcome is a classifier being able to deal with missing features.
The results in \autoref{fig:feat_imp} show that, unlike input perturbation, feature dropout  does not vastly overestimate features' importance. With feature dropout, it becomes apparent that only very few features  actually contain unique information, affecting accuracy when left out: the destination port and the source port.

@@

@@

Feature dropout might constitute a building block for methods based on Shapley values \cite{shapley_value_1953} like KernelSHAP \cite{lundberg_unified_2017}. 
%Due to space limitations, we leave this for future work.
\subsubsection{Mutual Information}
\begin{figure}
\includegraphics[width=\columnwidth]{./plots/mutinfo_example.pdf}
\caption{Two distributions yielding an identical accuracy drop.}
\label{fig:mutinfo_example}
\end{figure}
Input perturbation and feature dropout mainly address feature importance.
For example, assuming that the test dataset is representative for production use, for feature importance it is reasonable to equate the distribution for perturbing a feature with the feature distribution itself. However, when evaluating feature sensitivity, e.g. for analyzing potential for adversarial samples, the attacker is not limited by this distribution and frequently is able to choose arbitrary values in the feature space.

\begin{figure*}
\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2015.pdf}
\caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
\label{fig:feat_imp}
\end{figure*}
Furthermore, let $f(A,x)$  denote the joint probability
%\todo{Max: Ist joint probability nicht immer wenn man mehrere Ereignisse hat? Ich hätte hier probability density function geschrieben...}
%AH. changed notation and wording
for classification as attack and a feature value $x$. Accuracy then is $\int_{\mathbb R} f(A,x) dx$ for attacks and $1-\int_{\mathbb R} f(A,x) dx$ for benign traffic.
\autoref{fig:mutinfo_example} shows two different distributions
yielding the same accuracy, but clearly the right-side distribution has a larger influence on the prediction, as in the right-side case the prediction deterministically depends on the feature value.
To capture such dependencies, we propose to use mutual information %instead of accuracy drop
to determine feature sensitivity.
Mutual Information between two random variables $X,Y$ is defined as
\begin{equation}
I_{X,Y} = \mathbb E \left\{ \log\left(\frac{f_{X,Y}(x,y)}{f_X(x)f_Y(y)}\right) \right\},
\end{equation}
with $f_X(x), f_Y(y)$ and $f_{X,Y}(x,y)$ denoting the distribution of $X,Y$ and their joint distribution, respectively. To obtain feature sensitivity, we compute mutual information between an input variable and the prediction output for one flow at one particular time step, averaging over the result for the test set.

@@

The weights method is able to reveal features with a very low importance to a certain degree, but disagrees with the other methods to a large extent. Less anticipated, however, also input perturbation does not exhibit a considerable correlation with feature dropout. Considering its functioning of completely removing individual features, feature dropout is the most reliable method for evaluating importance for removing features. It is remarkable that none of the other methods is able to depict the distinct peak of importance for the destination port.

@@

@@

Due to the use of a 5-tuple flow key, port numbers and the protocol identifier are constant for all packets in one flow.
Hence, we can consider the \gls{rnn} a regular classifier and reuse 
%which allows us to use
\glspl{pdp}, which have been proposed for non-recurrent classification methods, by plotting the \gls{rnn}'s flow prediction outcome over one of these  features.
The results (omitted for brevity) show that for some attack types the port numbers play an important role. When looking at the \gls{pdp} for benign traffic samples, %(non-attack traffic)
it becomes apparent that traffic destined to a high destination port is generally indicative of an attack. We argue that this is because most services that regular users use have low port numbers.

@@

%As apparent from the above discussion, trying to explain how packet features influence the model's predictions, poses various difficulties. As a first step to explaining our model's predictions, we investigate how important features at different time steps are for the model's decisions.
Intuitively, features at the beginning of a flow should be the most important while the classifier's predictions should not vary significantly anymore, as soon as it has come to a decision.
\begin{figure}
\includegraphics[width=\columnwidth]{{"./plots/plot/flows_pred_plots2_outcomes_0_3.pickle_15_All_samples"}.pdf}
\caption{Classifier confidence per time step for CIC-IDS-2017. % and first and third quartiles.
For the majority of attack types, confidence increases in the first few steps and then stays almost constant at 1.}
\label{fig:confidence}
\end{figure}
To evaluate this hypothesis, \autoref{fig:confidence} depicts the classifier's prediction confidence for each time step, along with the number of samples having at least this length, which were used for evaluating the figure. % for each attack class.
%averaging over all flows. %AH. there's no averaging happening
% and also show the first and third quartiles.
%\autoref{fig:confidence} shows that %this hypothesis is correct for most attack classes.
While at the first couple of packets the confidence is not very high, towards the end of the flow it reaches values close to 1 and stays there. Hence, not only is the classifier able to make a reasonable classification after just a few packets, the figure also suggests that indeed later packets have a negligible influence on the prediction.

@@

\begin{figure}
%AH. I think the 'Firewall off' plot is not a good example as it shows only 4 packets
\includegraphics[width=\columnwidth]{{"./plots/plot_features/flows_pred_plots2_outcomes_0_3.pickle_2_Brute_Force-SSH-Patator"}.pdf}
\caption{\gls{iat} and packet length for SSH brute-force attacks in CIC-IDS-2017.}% The solid line shows the median value of the feature for all flows of a specific attack type. The shaded region are the first and third quartile.} % think legend is enough for unterstanding
\label{fig:plot_features}
\end{figure}
Finally, we investigated whether our classification task involves recognizing complex patterns in the feature space.
As example, \autoref{fig:plot_features} shows that
%Looking at the plots of the features that vary over time (\autoref{fig:plot_features}) we see that certain
attack types indeed have a characteristic pattern in which they send packets by which they are easily recognizable.
Other attack families similarly show characteristic patterns.
%We want to determine how the pattern of packet sizes and \glspl{iat} is significant for the classification. \autoref{fig:plot_features} clearly shows that some attack types have very characteristic patterns of packet sizes and \glspl{iat}.

\section{Defenses}
We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
\subsection{Reducing Features}
The most obvious defense is to simply omit features an attacker can manipulate. We try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all manipulable features: packet size and \gls{iat}
\item Leaving them out only in the direction from the attacker to the victim. This, however, does not prevent adversarial samples for botnets, for which the attacker has control over both sides.
\end{itemize}

@@

\subsection{Adversarial Training}
As alternative to omitting manipulable features, we attempted to make the classifier more robust against adversarial samples by augmenting the training set by adversarial flows generated using  \gls{cw}, labeled as additional attacks.
For this purpose, we added one adversarial sample per attack sample to the training set. Since \gls{cw} is deterministic, this is the maximum number of adversarial samples possible. With this augmented training set we then alternated retraining of the neural network and optimization of the adversarial samples. % using gradient descent.

@@

For both datasets, after a small number of epochs, it was no longer possible to create a significant number of adversarial samples for most attack categories. %For the datasets we considered, adversarial training left classification performance unaffected.
%\todo{Max: Insert results that show that accuracy on normal data is still ok after adversarial training.}
%\autoref{tab:adv_training_performance} depicts classification on non-adversarial data after 50 epochs of adversarial training.
After 50 epochs of training, accuracy is essentially identical to the results presented in \autoref{tab:performance_results}. Recall and Youden increased and precision slightly decreased. However, this is due to the higher proportion of attack samples in training data and, hence, expected.
We conclude that adversarial training is effective for reducing the attack surface for adversarial attacks in our scenario.

@@

We have implemented a recurrent classifier based on LSTMs to detect network attacks, %While attack detection performance is not necessarily better than that of non-recurrent approaches,
which can already detect attacks before they are over. Furthermore, the recurrent approach allows us to inspect the influence of single packets on the detection performance and shows which packets are \textit{characteristic} for attacks.
Even though the interpretation of \glspl{rnn} poses several difficulties, we have demonstrated methods for gaining insights into the model's functioning.

We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified. Newly developed \textit{feature sensitivity} methods can show which features can easily be manipulated by an adversary. We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.
Deploying an adversarial training procedure, we could significantly reduce the adversarial threat.

@@

\begin{figure}
\centering
\includegraphics[width=0.98\columnwidth]{./plots/ars/ars_15.pdf}
\caption{\gls{ars} during adversarial training for UNSW-NB15.}
\label{fig:avdistance}
\end{figure}

>>>>

We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. For a large-enough network, we do not expect these architectural parameters to have a severe impact on classification accuracy, so we chose these parameters to obtain a good performance while keeping training duration at an acceptable level.

As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e. forward or reverse path) and all TCP flags (0 if the flow is not TCP).
% \begin{itemize*}
% \item source port
% \item destination port
% \item protocol identifier
% \item packet sizes
% \item \glspl{iat} of packets
% \item packet directions (if packet goes there or back)
% \item all TCP flags (0 if the flow is not TCP)
% \end{itemize*}.
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behavior \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the other features vary.
During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization to transform feature values to the range appropriate for neural network training. We ensured that our classifiers do not suffer from overfitting using a train/test split of 2:1.

For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks. Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
\autoref{tab:performance_results} shows the achieved classification performance when evalutuating metrics per packet and per flow and includes performance results for an MLP classifier from~\cite{bachl_walling_2019} for comparison.
% \autoref{tab:performance_results} shows that
As depicted, our \gls{rnn}-based classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}. However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before the attack flows terminate.

\begin{table}
% Evaluated using
% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
% on Dec. 4th
\caption{Performance metrics per packet and per flow. MLP values from \cite{bachl_walling_2019} are presented for comparison.} \label{tab:performance_results}
\centering
\begin{tabular}{l r r r r r r} \toprule
& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
	&	Packet	&	Flow	& MLP &	Packet	&	Flow & MLP	\\	\midrule
Accuracy	&	99.1\%	&	99.7\%	&	99.8\%	&	99.5\%	&	98.3\%	&	98.9\%	\\
Precision	&	97.0\%	&	99.7\%	&	99.9\%	&	83.4\%	&	78.6\%	&	84.5\%	\\
Recall	&	97.8\%	&	99.1\%	&	99.2\%	&	87.6\%	&	72.6\%	&	82.9\%	\\
F1	&	97.4\%	&	99.4\%	&	99.5\%	&	85.5\%	&	75.5\%	&	83.7\%	\\
Youden's J	&	97.2\%	&	99.0\%	&	99.1\%	&	87.3\%	&	71.9\%	&	82.3\%	\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{figure}[b]
\includegraphics[width=\columnwidth]{./plots/adv_comparison/tradeoff.pdf}
\caption{Performance of \gls{cw} with different $\kappa$ values.}
\label{fig:adv_cw}
\end{figure}

@@

@@

\subsection{Feature Importance Metrics}
\begin{table}
\caption{Accuracy drop for:}
\label{tab:feat_importance}
\centering
\subfloat[Input perturbation\label{fig:featRandom}
]{
\begin{tabular}{l r}
\toprule
Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
Protocol & 0.207 \\
Packet Length & 0.165 \\
SYN Flag & 0.099 \\
ACK Flag & 0.084 \\
Direction & 0.073 \\
Destination port & 0.071 \\
Source port & 0.060 \\
RST Flag & 0.057 \\
PSH Flag & 0.056 \\
Interarrival time & 0.024 \\
FIN Flag & 0.012 \\
URG Flag & 0 \\
ECE Flag & 0 \\
CWR Flag & 0 \\
NS Flag & 0 \\
\bottomrule
\end{tabular}
}{}
\subfloat[Feature dropout\label{fig:featDropout}
]{
\begin{tabular}{l r}
\toprule
Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
Destination port & 0.025 \\
Source port & 0.003 \\
RST Flag & 0.001 \\
ACK Flag & 0.001 \\
Protocol & 0.001 \\
Packet Length & 0.001 \\
Direction & 0.001 \\
SYN Flag & 0 \\
Interarrival time & 0 \\
FIN Flag & 0 \\
ECE Flag & 0 \\
URG Flag & 0 \\
CWR Flag & 0 \\
NS Flag & 0 \\
PSH Flag & 0 \\
\bottomrule
\end{tabular}
}{}
\end{table}
As a first step to understanding the neural network's decisions, we  estimate how important individual features are for the model's predictions.
When investigating an \gls{ml} classifier, it is natural to ask which inputs have a large influence on the classifier's prediction. We feel the need to distinguish metrics for this purpose based on their main aim:

@@

\subsubsection{Neural Network Weights}
In previous works~\cite{olden_accurate_2004,olden_illuminating_2002}, a simple method for determining feature importance in neural networks has been summing up neural network weights leading from a certain input to the prediction outcome. The weights method can be considered for both feature importance and feature sensitivity, however, clearly, especially in the case of complex network architectures, this method is likely to provide wrong results. Hence, we provide results for the weights method mainly for comparison. Note that an LSTM cell alone has four weights leading from one input to an output.

@@

An important implementation detail is that for each feature we add another input which is 1 if the feature is suppressed and 0 otherwise. This is necessary for the neural network to be able to distinguish between a feature missing and a feature genuinely being zero. The overall outcome is a classifier being able to deal with missing features.
The results in \autoref{tab:feat_importance} show that, unlike input perturbation, feature dropout  does not vastly overestimate features' importance. With feature dropout, it becomes apparent that only very few features  actually contain unique information, affecting accuracy when left out: the destination port and the source port.

@@

\begin{figure*}
\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2015.pdf}
\caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
\label{fig:feat_imp}
\end{figure*}

\begin{figure}[b]
\includegraphics[width=\columnwidth]{./plots/mutinfo_example.pdf}
\caption{Two distributions yielding an identical accuracy drop.}
\label{fig:mutinfo_example}
\end{figure}

@@

Feature dropout might constitute a building block for methods based on Shapley values \cite{shapley_value_1953} like KernelSHAP \cite{lundberg_unified_2017}. 
%Due to space limitations, we leave this for future work.
\subsubsection{Mutual Information}
Input perturbation and feature dropout mainly address feature importance.
For example, assuming that the test dataset is representative for production use, for feature importance it is reasonable to equate the distribution for perturbing a feature with the feature distribution itself. However, when evaluating feature sensitivity, e.g. for analyzing potential for adversarial samples, the attacker is not limited by this distribution and frequently is able to choose arbitrary values in the feature space.

Furthermore, we argue that accuracy drop depicts an importance measure which might be misleading for evaluating feature sensitivity.
To see this, let $f(A,x)$  denote the joint probability
%\todo{Max: Ist joint probability nicht immer wenn man mehrere Ereignisse hat? Ich hätte hier probability density function geschrieben...}
%AH. changed notation and wording
for classification as attack and a feature value $x$. Accuracy then is $\int_{\mathbb R} f(A,x) dx$ for attacks and $1-\int_{\mathbb R} f(A,x) dx$ for benign traffic.
\autoref{fig:mutinfo_example} shows two different distributions
yielding the same accuracy, but clearly the right-side distribution has a larger influence on the prediction, as in the right-side case the prediction deterministically depends on the feature value.

To capture such dependencies, we propose to use mutual information %instead of accuracy drop
to determine feature sensitivity.
Mutual Information between two random variables $X,Y$ is defined as
\begin{equation}
I_{X,Y} = \mathbb E \left\{ \log\left(\frac{f_{X,Y}(x,y)}{f_X(x)f_Y(y)}\right) \right\},
\end{equation}
with $f_X(x), f_Y(y)$ and $f_{X,Y}(x,y)$ denoting the distribution of $X,Y$ and their joint distribution, respectively. To obtain feature sensitivity, we compute mutual information between an input variable and the prediction output for one flow at one particular time step, averaging over the result for the test set.

@@

The weights method is able to reveal features with a very low importance to a certain degree, but disagrees with the other methods to a large extent. Less anticipated, however, also input perturbation does not exhibit a considerable correlation with feature dropout. Considering its functioning of completely removing individual features, feature dropout is the most reliable method for evaluating importance for removing features. It is remarkable that none of the other methods is able to depict the distinct peak of importance for the destination port visible for feature dropout in \autoref{fig:feat_imp} and \autoref{tab:feat_importance}.

@@

\begin{figure}
\includegraphics[width=\columnwidth]{{"../plots/plot_pdp/flows_pdp_outcomes_0_3.pickle_10_Normal"}.pdf}
\caption{PD plot for the source and destination port features.}
\label{fig:pdps}
\end{figure}

@@

Due to the use of a 5-tuple flow key, port numbers and the protocol identifier are constant for all packets in one flow.
Hence, we can consider the \gls{rnn} a regular classifier and reuse 
%which allows us to use
\glspl{pdp}, which have been proposed for non-recurrent classification methods, by plotting the \gls{rnn}'s flow prediction outcome over one of these  features.
The results show that for some attack types the port numbers play an important role. When looking at the \gls{pdp} for benign traffic samples in \autoref{fig:pdps}, %(non-attack traffic)
it becomes apparent that traffic destined to a high destination port is generally indicative of an attack. We argue that this is because most services that regular users use have low port numbers.

@@

%As apparent from the above discussion, trying to explain how packet features influence the model's predictions, poses various difficulties. As a first step to explaining our model's predictions, we investigate how important features at different time steps are for the model's decisions.
Intuitively, features at the beginning of a flow should be the most important while the classifier's predictions should not vary significantly anymore, as soon as it has come to a decision.
\begin{figure}
\includegraphics[width=\columnwidth]{{"./plots/plot/flows_pred_plots2_outcomes_0_3.pickle_15_All_samples"}.pdf}
\caption{Classifier confidence per time step for CIC-IDS-2017. % and first and third quartiles.
For the majority of attack types, confidence increases in the first few steps and then stays almost constant at 1.}
\label{fig:confidence}
\end{figure}

To evaluate this hypothesis, \autoref{fig:confidence} depicts the classifier's prediction confidence for each time step, along with the number of samples having at least this length, which were used for evaluating the figure. % for each attack class.
%averaging over all flows. %AH. there's no averaging happening
% and also show the first and third quartiles.
%\autoref{fig:confidence} shows that %this hypothesis is correct for most attack classes.
While at the first couple of packets the confidence is not very high, towards the end of the flow it reaches values close to 1 and stays there. Hence, not only is the classifier able to make a reasonable classification after just a few packets, the figure also suggests that indeed later packets have a negligible influence on the prediction.

@@

Finally, we investigated whether our classification task involves recognizing complex patterns in the feature space.
As example, \autoref{fig:plot_features} shows that
%Looking at the plots of the features that vary over time (\autoref{fig:plot_features}) we see that certain
attack types indeed have a characteristic pattern in which they send packets by which they are easily recognizable.
Other attack families similarly show characteristic patterns.
%We want to determine how the pattern of packet sizes and \glspl{iat} is significant for the classification. \autoref{fig:plot_features} clearly shows that some attack types have very characteristic patterns of packet sizes and \glspl{iat}.

\section{Defenses}
We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks. Several methods have been proposed to improve robustness in the context of computer vision~\cite{akhtar_threat_2018}. We chose the methods presented below, because they are simple and can be applied to our recurrent setting in a straight-forward fashion.
\subsection{Reducing Features}
The most obvious defense is to simply omit features an attacker can manipulate. We try two different approaches of this defense strategy:
\begin{itemize}[leftmargin=0.5cm] %[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all manipulable features, i.e. packet size and \gls{iat}, in both directions.
\item Leaving them out only in the direction from the attacker to the victim. This, however, does not prevent adversarial samples for botnets, for which the attacker has control over both sides.
\end{itemize}

@@

\subsection{Adversarial Training}
As alternative to omitting manipulable features, we attempted to make the classifier more robust against adversarial samples by augmenting the training set by adversarial flows generated using  \gls{cw}, labeled as additional attacks. This approach can be considered an adversarial training procedure, which is a common defense method in the related literature~\cite{akhtar_threat_2018,madry_towards_2018}.
We added one adversarial sample per attack sample to the training set. Since \gls{cw} is deterministic, this is the maximum number of adversarial samples possible. With this augmented training set we then alternated retraining of the neural network and optimization of the adversarial samples. % using gradient descent.

@@

\begin{figure}[t]
%AH. I think the 'Firewall off' plot is not a good example as it shows only 4 packets
\includegraphics[width=\columnwidth]{{"./plots/plot_features/flows_pred_plots2_outcomes_0_3.pickle_2_Brute_Force-SSH-Patator"}.pdf}
\caption{\gls{iat} and packet length for SSH brute-force attacks in CIC-IDS-2017.}% The solid line shows the median value of the feature for all flows of a specific attack type. The shaded region are the first and third quartile.} % think legend is enough for unterstanding
\label{fig:plot_features}
\end{figure}

\begin{figure}[b]
\centering
\includegraphics[width=0.98\columnwidth]{./plots/ars/ars_15.pdf}
\caption{\gls{ars} during adversarial training for UNSW-NB15.}
\label{fig:avdistance}
\end{figure}

For both datasets, after a small number of epochs, it was no longer possible to create a significant number of adversarial samples for most attack categories. %For the datasets we considered, adversarial training left classification performance unaffected.
%\todo{Max: Insert results that show that accuracy on normal data is still ok after adversarial training.}
%\autoref{tab:adv_training_performance} depicts classification on non-adversarial data after 50 epochs of adversarial training.
After 50 epochs of training, accuracy is essentially identical to the results presented in \autoref{tab:performance_results}. Recall and informedness increased and precision slightly decreased. However, this is due to the higher proportion of attack samples in training data and, hence, expected.
We conclude that adversarial training is effective for reducing the attack surface for adversarial attacks in our scenario.

@@

We have implemented a recurrent classifier based on LSTMs to detect network attacks, %While attack detection performance is not necessarily better than that of non-recurrent approaches,
which is able to  detect attacks already before they are over. The recurrent approach allowed us to inspect the influence of single packets on the detection performance and shows which packets are \textit{characteristic} for attacks.
Even though the interpretation of \glspl{rnn} poses several difficulties, we have demonstrated methods for gaining insights into the model's functioning.

We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified. We introduced \textit{feature sensitivity} methods to show which features can easily be manipulated by an adversary to cause a wrong classification. We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.
Deploying an adversarial training procedure, we could significantly reduce the adversarial threat.

@@
